\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{amsmath, amsfonts}
\usepackage{hyperref, graphicx}
\usepackage{tikz}

\DeclareMathOperator*{\argmin}{arg\,min}

\title{CSCI 145 Problem Set 2}
\author{} % TODO: Put your name here
\date{\today}

\begin{document}

\maketitle

\subsection*{Submission Instructions}

Please upload \textit{your} work by
\textbf{11:59pm Monday September 15, 2025.}
\begin{itemize}
\item You are encouraged to discuss ideas
and work with your classmates. However, you
\textbf{must acknowledge} your collaborators
at the top of each solution on which
you collaborated with others 
and you \textbf{must write} your solutions
independently.
\item Your solutions to theory questions must
be written legibly, or typeset in LaTeX or markdown.
If you would like to use LaTeX, you can import the source of this document (available from the course webpage) to Overleaf.
\item I recommend that you write your solutions to coding questions in a Jupyter notebook using Google Colab.
\item You should submit your solutions as a \textbf{single PDF} via the assignment on Gradescope.
\end{itemize}

\noindent
\textbf{Grading:} The point of the problem set is for \textit{you} to learn. To this end, I hope to disincentivize the use of LLMs by \textbf{not} grading your work for correctness. Instead, you will grade your own work by comparing it to my solutions. This self-grade is due the Friday \textit{after} the problem set is due, also on Gradescope.

\newpage
\section*{Problem 1: Gradient Descent on Quadratics}

Let's apply gradient descent to quadratic loss functions.
Recall the gradient descent update is
\begin{align}
    \mathbf{w}^{(t+1)} \gets 
    \mathbf{w}^{(t)}
    - \alpha \nabla \mathcal{L}(\mathbf{w}).
\end{align}

\subsection*{Part A: Closed-form Iterates}
Consider minimizing a one-dimensional quadratic loss
$\mathcal{L}(w) = \frac{a}{2} w^2 - b\, w + c$ with $a>0.$

Show that the error contracts geometrically:
\[
w^{(t)} - w^* = (1 - \alpha a)^t \,(w^{(0)} - w^*).
\]
For which $\alpha$ does this converge? What choice of $\alpha$ gives the fastest (one-step) convergence in 1D?

\textbf{Hint:} Find $w^*$, and write $b$ in terms of $w^*$.

\subsection*{Part B: Multi-dimensional Extension}
Let $\mathcal{L}(\mathbf{w}) = \tfrac12 \mathbf{w}^\top \mathbf{A} \mathbf{w} - \mathbf{b}^\top \mathbf{w} + c$ with $\mathbf{A} \in \mathbb{R}^{d\times d}$ symmetric positive definite (all eigenvalues are strictly positive).
Show that for the gradient descent update 
$$
\|\mathbf{w}^{(t)}-\mathbf{w}^*\|_2 \le \rho^t \|\mathbf{w}^{(0)}-\mathbf{w}^*\|_2
\quad\text{where}\quad
\rho = \max_i |1-\alpha \lambda_i(\mathbf{A})|.
$$
Deduce that convergence holds iff $0<\alpha<2/\lambda_{\max}(\mathbf{A})$.

\textbf{Hint:} Follow the same approach as before. Along the way, use the inequality that $\| \mathbf{M v} \|_2 \leq \| \mathbf{M} \|_2 \|\mathbf{v}\|_2$, for matrix $\mathbf{M}$ and vector $\mathbf{v}$, where $\| \mathbf{M} \|_2 = \lambda_{\text{max}} (\mathbf{M})$ is the \textit{spectral} norm of a matrix.

\subsection*{Part C: Least Squares Specialization}
For the MSE loss $\mathcal{L}(\mathbf{w})=\tfrac1{2}\|\mathbf{X}\mathbf{w}-\mathbf{y}\|_2^2$, identify $\mathbf{A}$ and $\mathbf{b}$ and express $\lambda_{\max}(\mathbf{A})$ in terms of $\mathbf{X}$. What step-size bound in terms of the singular value decomposition of $\mathbf{X}$ guarantees convergence?

\subsection*{Part D: Empirical Checks}

Load a regression dataset of your choice.
With random initializations, repeatedly run gradient descent for $100$ epochs with various learning rates.
In particular, choose $\alpha$ at evenly spaced intervals from $0$ to twice the bound you found in the prior part.
Plot the loss by these $\alpha$.
What do you notice?

%\input{solutions/solution3_1}

\end{document}