\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{amsmath, amsfonts}
\usepackage{hyperref, graphicx}

\DeclareMathOperator*{\argmin}{arg\,min}

\title{CSCI 145 Problem Set 1}
\author{} % TODO: Put your name here
\date{\today}

\begin{document}

\maketitle

\subsection*{Submission Instructions}

Please upload \textit{your} work by
\textbf{5pm Friday September 19, 2025.}
\begin{itemize}
\item You are encouraged to discuss ideas
and work with your classmates. However, you
\textbf{must acknowledge} your collaborators
at the top of each solution on which
you collaborated with others 
and you \textbf{must write} your solutions
independently.
\item Your solutions to theory questions must
be written legibly, or typeset in LaTeX or markdown.
If you would like to use LaTeX, you can import the source of this document (available from the course webpage) to Overleaf.
\item I recommend that you write your solutions to coding questions in a Jupyter notebook using Google Colab.
\item You should submit your solutions as a \textbf{single PDF} via the assignment on Gradescope.
\end{itemize}

\noindent
\textbf{Grading:} The point of the problem set is for \textit{you} to learn. To this end, I hope to disincentivize the use of LLMs by \textbf{not} grading your work for correctness. Instead, you will grade your own work by comparing it to my solutions. This self-grade is due the Monday \textit{after} the problem set is due, also on Gradescope.

\newpage \section*{Problem 1: Fast Matrix-Vector Multiplication}

Let $d$ and $q$ be non-negative integers.
Consider matrix $\mathbf{A} \in \mathbb{R}^{d \times d}$, and vector $\mathbf{x} \in \mathbb{R}^d$.
The goal in this problem is to compute $\mathbf{A}^q \mathbf{x}$ quickly with increasing levels of sophistication.

\subsection*{Part A: Naive Attempt}

Describe a method to compute $\mathbf{A}^q \mathbf{x}$ in time $O(q \cdot d^3)$.

\subsection*{Part B: Strategic Matrix Multiplication}

For convenience, suppose that $q$ is a power of 2.
Describe a method to compute $\mathbf{A}^q \mathbf{x}$ in time $O(\log_2 q \cdot d^3)$.

\subsection*{Part C: Vector Multiplication}

Describe a method to compute $\mathbf{A}^q \mathbf{x}$ in time $O(q \cdot d^2)$.

\subsection*{Part D: Eigendecomposition}

Suppose that $c^q$ can be computed in constant time, for any real number $c \in \mathbb{R}$.
Describe a method to compute $\mathbf{A}^q \mathbf{x}$ in time $O(d^3)$.

\subsection*{Part E: Empirical Checks}

Initialize a vector $\mathbf{x}$ and \textit{symmetric} matrix $\mathbf{A}$, where each entry is drawn independently from a Gaussian distribution.
Implement your various strategies in python, and plot their runtimes for $d=10$ and $q \in \{2, 4, 8, 16, 32, 64, 128, 256\}$.

%\input{solutions/solution1_1}

\end{document}